The transiting dust clumps in the evolved disk of the Sun-like UXOr RZ~Psc

### 2016-08-31 ms.tex
submission to rsos

What separates RZ~Psc from other UXOrs (and dippers) is i) the spectral type is K0V
rather than Herbig Ae/Be for UXOrs and M-type for dippers, ii) the occulting dust lies
well beyond the sublimation radius, and iii) the star is not associated with a
star-forming region so is inferred to be a few tens of millions of years old. The dust
distance has been inferred from the speed of ingress of dimming events, which was
previously estimated as about 0.6au (for circular orbits
\cite{2013A&A...553L...1D}). Corroborating evidence comes from the $\sim$500K temperature
of the dust seen in the mid-IR, which places it near 0.4-0.7au (depending on optical
depth) and therefore at a location consistent with the occulting dust
\cite{2013A&A...553L...1D}. The distance to RZ~Psc is unknown, but as an apparently
isolated star that shows Li absorption the age has been estimated as a few tens of Myr,
and therefore beyond the age at which a gas-rich disk should exist
\cite{2010A&A...524A...8G,2014A&A...563A.139P}. Further distinguishing features are that
the duration of the dimming events is relatively short compared to other UXOrs, a few
days rather than a few weeks, and that no near-IR excess or accretion signatures are seen
\cite{2014A&A...563A.139P}, so interpretations related to accretion of disk material onto
the star (e.g. \cite{1999AJ....118.1043H,1999A&A...349..619B,2016arXiv160503985B}) are
unlikely.

Thus, the potentially compelling and unique aspect for RZ~Psc is that we are observing
dimming events from dust in a main-sequence planetary system that resides at about
0.5au. This dust is also seen in thermal emission, so deriving joint constraints on the
dust properties and structure may be possible. As argued by \cite{2013A&A...553L...1D} a
picture is emerging in which RZ~Psc is surrounded by a massive young version of our own
asteroid belt, in which planetesimals are continually being destroyed. These collisions
generate the large collective surface area of small dust that emits strongly in the
mid-IR, and the system geometry means that this dust also sometimes passes in front of
the star.

@@

To study the temporal variability of RZ~Psc we use two seasons of public data from the
Wide-Angle Search for Planets (WASP \cite{2006PASP..118.1407P}), and nine seasons of data
from the Kilodegree Extremely Little Telescope North (KELT-North
\cite{2007PASP..119..923P}. We also collected, but ultimately did not use, photometric
observations of RZ~Psc from a wide variety of other sources
(\cite{1994AJ....108.1906H,1973IBVS..783....1K,1980PZ.....21..310K,1985PZ.....22..181Z,1991Afz....34..333K,1997AcA....47..467P,2014Ap.....57..491P},
the Catalina Sky Survey, the American Association of Variable Star Observers, the All-Sky
Automated Survey). Most of these other data have been previously published and their
collation here is largely for the benefit of others. This compilation contains nearly all
data for which tables have been published, but aside from the Harvard plate photometry
published by \cite{1999A&AS..140..293G} we have not sought other photometry so the light
curve remains incomplete.\footnote{Normalised photometry is available at
  \href{https://github.com/drgmk/rzpsc}{https://github.com/drgmk/rzpsc}}

@@

% plot_hist_sim.py
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/hist_single.ps}
    \caption{Power at a given period from the iterative event finding. The solid line
      shows the power from the data, and is the same in each panel. The dashed lines show
      the mean and $\pm$1$\sigma$ power from simulated dimming events, which from top to
      bottom are: random, periodic between 64 and 75 days with 3 repeats, and periodic at
      64 days with 10 repeats.}\label{fig:hist}
  \end{center}
\end{figure}

@@

The simplest case is a star of uniform surface brightness that is occulted by an
optically thick screen that covers the star from $x=-1$ to $u$ (i.e. the units of length
are now $R_\star$). Then $I=1$ and $\tau(x-x_{\rm cl})$ is a step function at $u$ and the
% observed stellar flux is
% \begin{equation}
%   F = 2 \int_{u}^{1} \sqrt{1-x^2} dx
% \end{equation}
% \cite{2006ApJ...644..510W}. The 
fraction of the total stellar flux ($F_\star = \pi$) seen is
\begin{equation}
  f = F/F_\star = (cos^{-1}[u] - u\sqrt{1-u^2})/\pi
\end{equation}
\cite{2006ApJ...644..510W}, where $f$ has the same units as our normalised light
curve. If the curtain is not completely optically thick then the fraction is instead
\begin{equation}
 f = F/F_\star + (1-F/F_\star)(1-\tau)
\end{equation}

@@

% gradients.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/gradients.eps}
    \caption{Gradient and minimum flux measured from individual nights'
      observations. Open circles have gradients significantly different from zero, and a
      symbol size proportional to the inverse of the gradient uncertainty. Dots are
      consistent with zero slope. The lines show the gradients implied by the velocities
      for circular orbits at 0.3, 1, 10, and 100au.}\label{fig:grad}
  \end{center}
\end{figure}

@@

% gradients.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/sma.eps}
    \caption{Semi-major axes estimated from the light curve gradients in Figure
      \ref{fig:grad}. Open circles have gradients significantly different from zero, and
      a symbol size proportional to the inverse of the gradient uncertainty. Dots are
      consistent with zero slope. The stellar radius (grey dashed line) has been
      estimated as Solar. Because gradients may have lower minimum fluxes they can move
      down along lines parallel to the solid lines. Points near or above the blue line
      are consistent with zero light curve gradient.}\label{fig:sma}
  \end{center}
\end{figure}

In Figure \ref{fig:sma} we have converted the gradients into semi-major axes using
equation (\ref{eq:acl}). The dashed line shows the stellar radius, estimated to be
approximately Solar based on an age of 25Myr \cite{2013AstL...39..776P} and an effective
temperature of 5350K \cite{2014A&A...563A.139P} using \cite{2000A&A...358..593S}
isochrones. This Figure also includes solid lines of constant gradient, computed using
equation (\ref{eq:acl}) and $f_{\rm min} = 1-\tau$.

@@

The aim of this section was to estimate the location of the bodies that cause the dimming
events towards RZ~Psc. Using ten years of WASP and KELT-North data, we investigated both the
periodicity of possible repeat events, and the light curve gradients. While we found
evidence for a 60-70 day period, this signal is relatively weak. This period would place
the clumps near 0.3au, similar to the distance inferred for the dust seen as an IR excess
(though this location is also uncertain). The gradient analysis places loose constraints
on the clump orbits ($<$10au), so is consistent with a scenario where the clumps have
semi-major axes near 0.3au. Thus, there is no reason to disfavour the model proposed by
\cite{2013A&A...553L...1D}, that the dimming events are associated with clumps being
created by planetesimal collisions within an asteroid belt analogue.

@@

% rcl-a.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/rcl-a.eps}
    \caption{Clump properties assuming circular orbits for a range of dimming event
      durations (as labelled). The brown region marks the approximate location of the
      RZ~Psc asteroid belt (or the inner edge of a more extended disk). The blue shaded
      region shows a typical Herbig Ae/Be inner disk edge near 1au. The PINK? line shows
      where a clump has an azimuthal extent similar to the scale height of a typical
      gas-rich disk.}\label{fig:rcla}
  \end{center}
\end{figure}

@@

% sedcomp.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/spcomp.eps}
    \caption{The SED of RZ~Psc in comparison with other young disk-hosting stars. The
      triangles are AKARI and IRAS upper limits for RZ~Psc. Measurements we made at
      different times, so the apparent discrepancy between the WISE detection at 22$\mu$m
      and the AKARI upper limit at 18$\mu$m is probable indicator of IR
      variability.}\label{fig:spcomp}
  \end{center}
\end{figure}

@@

% export from cartoon.key, cropped with pdfcrop
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.5\textwidth]{figs/cartoon-crop.ps}
    \caption{Cartoon showing possible origins of dippers and UXOrs, and why Sun-like
      stars may only rarely show analogous behaviour. In each case the star, magnetic
      dipole, and rotation axis are shown at the left, with possible disk structures
      viewed edge-on to the right. (\emph{top}) low-mass stars (dippers) are occulted by
      co-rotating material that is accreting on to the star, and the dust sublimation
      radius is interior to corotation \cite{2016arXiv160503985B}. (\emph{middle})
      Sun-like stars are rarely seen as dippers or UXOrs because i) dust sublimates
      outside corotation (represented by the grey accretion column in the upper half
      \cite{2016arXiv160503985B}), or ii) material lifted by turbulence is shadowed by
      the outer disk (spiral in the lower half). (\emph{bottom}) Herbig Ae/Be stars
      (UXOrs) are occulted by turbulence that appears above self-shadowed disks
      \cite{2003ApJ...594L..47D}. In this type of (primordial) scenario RZ~Psc is
      occulted by turbulence in an evolved disk with little flaring (i.e. dust has
      settled towards the disk midplane).}\label{fig:cartoon}
  \end{center}
\end{figure}

@@

From the perspective of dippers, which so far have spectral types later than K5,
\cite{2016arXiv160503985B} explain their tendency to be low-mass stars as a consequence
of the relation between the magnetospheric truncation, corotation, and sublimation
radii. The periodicity of dippers suggests that the occulting material is near the
corotation radius, and is therefore near the base of any accretion columns. For low-mass
stars the dust temperature at this distance is cool enough that the columns contain
significant dust mass and hence the dipping phenomenon is seen (top panel of Figure
\ref{fig:cartoon}. For earlier type stars the sublimation radius is outside the
corotation radius, so any dust in the accretion columns has sublimated and no dipping is
seen (grey accretion column in the middle panel of Figure \ref{fig:cartoon}).

@@

From the perspective of UXOrs, a possible explanation for their tendency to be B and
A-type stars is that the specifics of self-shadowing are different for Sun-like stars
\cite{2003ApJ...594L..47D}. The reason is unclear, but it seems either that
self-shadowing is simply rarer, or that the nature is different \cite[i.e. shadowing by
the inner disk regions, rather than by a puffed-up inner rim,
see]{2004A&A...421.1075D,2007prpl.conf..555D}

@@

Data and code used in the preparation of this contribution can be found online at
\href{https://github.com/drgmk/rzpsc}{https://github.com/drgmk/rzpsc}.

We declare that we have no competing interests.

GMK initiated the project, collated data, did most of the analysis, and wrote the
paper. MAK contributed early ideas and the iterative event finding method. KGS provided
independent SED models to test for a UV excess. As members of the KELT collaboration, JP,
JER, RJS, and KGS acquired the time-series photometry. All co-authors provided input on
the style and content of the manuscript.

We thank Joachim G\"urtler for sharing photometry from the Sonneberg and Harvard Plates
in a palatable form, AAVSO observers for monitoring RZ~Psc, and Simon Hodgkin and Mark
Wyatt for useful discussions. This paper makes use of data from the DR1 of the WASP data
\cite{2010A&A...520L..10B} as provided by the WASP consortium, and the computing and
storage facilities at the CERIT Scientific Cloud, reg. no. CZ.1.05/3.2.00/08.0144 which
is operated by Masaryk University, Czech Republic.

GMK is supported by the Royal Society as a Royal Society University Research Fellow. MAK
is employed as a lecturer at the Leiden Observatory. JP is employed as a lecturer at
Lehigh University. JER is funded as a Future Faculty Leaders Fellow at the
Harvard-Smithsonian Center for Astrophysics. RS is employed as an Instrumentation
Scientist for the Las Cumbres Observatory Global Telescope Network. KGS is funded as a
lecturer at Vanderbilt University.

>>>>

What separates RZ~Psc from other UXOrs (and dippers) is i) the spectral type is K0V
rather than Herbig Ae/Be for UXOrs and late K to M-type for dippers, ii) the occulting
dust lies well beyond the sublimation radius, and iii) the star is not associated with a
star-forming region so is inferred to be a few tens of millions of years old. The dust
distance has been inferred from the speed of ingress of dimming events, which was
previously estimated as about 0.6au (for circular orbits
\cite{2013A&A...553L...1D}). Corroborating evidence comes from the $\sim$500K temperature
of the dust seen in the mid-IR, which places it near 0.4-0.7au (depending on optical
depth) and therefore at a location consistent with the occulting dust
\cite{2013A&A...553L...1D}. The distance to RZ~Psc is unknown, but as an apparently
isolated star that shows Li absorption the age has been estimated as a few tens of Myr,
and therefore beyond the age at which a gas-rich disk should exist
\cite{2010A&A...524A...8G,2014A&A...563A.139P}. Further distinguishing features are that
the duration of the dimming events is relatively short compared to other UXOrs, a few
days rather than a few weeks, and that no near-IR excess or accretion signatures are seen
\cite{2014A&A...563A.139P}, so interpretations related to accretion of disk material onto
the star (e.g. \cite{1999AJ....118.1043H,1999A&A...349..619B,2016arXiv160503985B}) are
unlikely.

Thus, the potentially compelling and unique aspect for RZ~Psc is that we are observing
dimming events from dust in a main-sequence planetary system that resides at about
0.5au. This dust is also seen in thermal emission, so deriving joint constraints on the
dust properties and structure may be possible. As argued by de~Wit et
al. \cite{2013A&A...553L...1D} a picture is emerging in which RZ~Psc is surrounded by a
massive young version of our own asteroid belt, in which planetesimals are continually
being destroyed. These collisions generate the large collective surface area of small
dust that emits strongly in the mid-IR, and the system geometry means that this dust also
sometimes passes in front of the star.

@@

To study the temporal variability of RZ~Psc we use two seasons of public data from the
Wide-Angle Search for Planets (WASP \cite{2006PASP..118.1407P}), and nine seasons of data
from the Kilodegree Extremely Little Telescope North (KELT-North
\cite{2007PASP..119..923P}). We also collected, but ultimately did not use, photometric
observations of RZ~Psc from a wide variety of other sources
(\cite{1994AJ....108.1906H,1973IBVS..783....1K,1980PZ.....21..310K,1985PZ.....22..181Z,1991Afz....34..333K,1997AcA....47..467P,2014Ap.....57..491P},
the Catalina Sky Survey, the American Association of Variable Star Observers, the All-Sky
Automated Survey). Most of these other data have been previously published and their
collation here is largely for the benefit of others. This compilation contains nearly all
data for which tables have been published, but aside from the Harvard plate photometry
published by G\:urtler et al. \cite{1999A&AS..140..293G} we have not sought other
photometry so the light curve remains incomplete.\footnote{Normalised photometry is
  available at \href{https://github.com/drgmk/rzpsc}{https://github.com/drgmk/rzpsc}}

@@

% plot_hist_sim.py
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/hist_single.ps}
    \caption{Power at a given period from the iterative event finding. The solid line
      shows the power from the data, and is the same in each panel. The dashed lines show
      the mean and $\pm$1$\sigma$ power from simulated dimming events, which from top to
      bottom are: random, periodic between 64 and 75 days with 3 repeats, and periodic at
      64 days with 10 repeats.}\label{fig:hist}
  \end{center}
\end{figure}

@@

The simplest case is a star of uniform surface brightness that is occulted by an
optically thick screen that covers the star from $x=-1$ to $u$ (i.e. the units of length
are now $R_\star$). Then $I=1$ and $\tau(x-x_{\rm cl})$ is a step function at $u$ and the
% observed stellar flux is
% \begin{equation}
%   F = 2 \int_{u}^{1} \sqrt{1-x^2} dx
% \end{equation}
% \cite{2006ApJ...644..510W}. The 
fraction of the total stellar flux ($F_\star = \pi$) seen is \cite{2006ApJ...644..510W}
\begin{equation}
  f = F/F_\star = (cos^{-1}[u] - u\sqrt{1-u^2})/\pi \, ,
\end{equation}
where $f$ has the same units as our normalised light curve. If the curtain is not
completely optically thick then the fraction is instead
\begin{equation}
 f = F/F_\star + (1-F/F_\star)(1-\tau)
\end{equation}

@@

% gradients.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/gradients.eps}
    \caption{Gradient and minimum flux measured from individual nights'
      observations. Open circles have gradients significantly different from zero, and a
      symbol size proportional to the inverse of the gradient uncertainty. Dots are
      consistent with zero slope. The lines show the gradients implied by the velocities
      for circular orbits at 0.3, 1, 10, and 100au.}\label{fig:grad}
  \end{center}
\end{figure}

@@

% gradients.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/sma.eps}
    \caption{Semi-major axes estimated from the light curve gradients in Figure
      \ref{fig:grad}. Open circles have gradients significantly different from zero, and
      a symbol size proportional to the inverse of the gradient uncertainty. Dots are
      consistent with zero slope. The stellar radius (grey dashed line) has been
      estimated as Solar. Because gradients may have lower minimum fluxes they can move
      down along lines parallel to the solid lines. Points near or above the blue line
      are consistent with zero light curve gradient.}\label{fig:sma}
  \end{center}
\end{figure}

In Figure \ref{fig:sma} we have converted the gradients into semi-major axes using
equation (\ref{eq:acl}). The dashed line shows the stellar radius, estimated to be
approximately Solar based on an age of 25Myr \cite{2013AstL...39..776P} and an effective
temperature of 5350K \cite{2014A&A...563A.139P} using Siess \cite{2000A&A...358..593S}
isochrones. This Figure also includes solid lines of constant gradient, computed using
equation (\ref{eq:acl}) and $f_{\rm min} = 1-\tau$.

@@

The aim of this section was to estimate the location of the bodies that cause the dimming
events towards RZ~Psc. Using ten years of WASP and KELT-North data, we investigated both
the periodicity of possible repeat events, and the light curve gradients. While we found
evidence for a 60-70 day period, this signal is relatively weak. This period would place
the clumps near 0.3au, similar to the distance inferred for the dust seen as an IR excess
(though this location is also uncertain). The gradient analysis places loose constraints
on the clump orbits ($<$10au), so is consistent with a scenario where the clumps have
semi-major axes near 0.3au. Thus, there is no reason to disfavour the model proposed by
de Wit et al. \cite{2013A&A...553L...1D}, that the dimming events are associated with
clumps being created by planetesimal collisions within an asteroid belt analogue.

@@

% rcl-a.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/rcl-a.eps}
    \caption{Clump properties assuming circular orbits for a range of dimming event
      durations (as labelled). The brown region marks the approximate location of the
      RZ~Psc asteroid belt (or the inner edge of a more extended disk). The blue shaded
      region shows a typical Herbig Ae/Be inner disk edge near 1au. The PINK? line shows
      where a clump has an azimuthal extent similar to the scale height of a typical
      gas-rich disk.}\label{fig:rcla}
  \end{center}
\end{figure}

@@

% sedcomp.pro
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/spcomp.eps}
    \caption{The SED of RZ~Psc in comparison with other young disk-hosting stars. The
      triangles are AKARI and IRAS upper limits for RZ~Psc. Measurements we made at
      different times, so the apparent discrepancy between the WISE detection at 22$\mu$m
      and the AKARI upper limit at 18$\mu$m is probable indicator of IR
      variability.}\label{fig:spcomp}
  \end{center}
\end{figure}

@@

% export from cartoon.key, cropped with pdfcrop
\begin{figure}
  \begin{center}
    \hspace{-0.5cm} \includegraphics[width=0.7\textwidth]{figs/cartoon-crop.ps}
    \caption{Cartoon showing possible origins of dippers and UXOrs, and why Sun-like
      stars may only rarely show analogous behaviour. In each case the star, magnetic
      dipole, and rotation axis are shown at the left, with possible disk structures
      viewed edge-on to the right. (\emph{top}) low-mass stars (dippers) are occulted by
      co-rotating material that is accreting on to the star, and the dust sublimation
      radius is interior to corotation \cite{2016arXiv160503985B}. (\emph{middle})
      Sun-like stars are rarely seen as dippers or UXOrs because i) dust sublimates
      outside corotation (represented by the grey accretion column in the upper half
      \cite{2016arXiv160503985B}), or ii) material lifted by turbulence is shadowed by
      the outer disk (spiral in the lower half). (\emph{bottom}) Herbig Ae/Be stars
      (UXOrs) are occulted by turbulence that appears above self-shadowed disks
      \cite{2003ApJ...594L..47D}. In this type of (primordial) scenario RZ~Psc is
      occulted by turbulence in an evolved disk with little flaring (i.e. dust has
      settled towards the disk midplane).}\label{fig:cartoon}
  \end{center}
\end{figure}

@@

From the perspective of dippers, which so far have spectral types later than K5, Bodman
et al.  \cite{2016arXiv160503985B} explain their tendency to be low-mass stars as a
consequence of the relation between the magnetospheric truncation, corotation, and
sublimation radii. The periodicity of dippers suggests that the occulting material is
near the corotation radius, and is therefore near the base of any accretion columns. For
low-mass stars the dust temperature at this distance is cool enough that the columns
contain significant dust mass and hence the dipping phenomenon is seen (top panel of
Figure \ref{fig:cartoon}. For earlier type stars the sublimation radius is outside the
corotation radius, so any dust in the accretion columns has sublimated and no dipping is
seen (grey accretion column in the middle panel of Figure \ref{fig:cartoon}).

@@

From the perspective of UXOrs, a possible explanation for their tendency to be B and
A-type stars is that the specifics of self-shadowing are different for Sun-like stars
\cite{2003ApJ...594L..47D}. The reason is unclear, but it seems either that
self-shadowing is simply rarer, or that the nature is different (i.e. shadowing by the
inner disk regions rather than by a puffed-up inner rim, see
\cite{2004A&A...421.1075D,2007prpl.conf..555D}).

@@

\textbf{Data and code:} Materials used in the preparation of this contribution can be
found online at \href{https://github.com/drgmk/rzpsc}{https://github.com/drgmk/rzpsc}.

\textbf{Competing interests:} We declare that we have no competing interests.

\textbf{Contributions:} GMK initiated the project, collated data, did most of the
analysis, and wrote the paper. MAK contributed to the time-series analysis and conceived
the iterative event finding method. KGS provided independent SED models to test for a UV
excess. As members of the KELT collaboration, JP, JER, RJS, and KGS acquired the
time-series photometry. All co-authors provided input on the style and content of the
manuscript.

\textbf{Acknowledgments:} We thank Joachim G\"urtler for sharing photometry from the
Sonneberg and Harvard Plates in a palatable form, AAVSO observers for monitoring RZ~Psc,
and Simon Hodgkin and Mark Wyatt for useful discussions. This paper makes use of data
from the DR1 of the WASP data \cite{2010A&A...520L..10B} as provided by the WASP
consortium, and the computing and storage facilities at the CERIT Scientific Cloud,
reg. no. CZ.1.05/3.2.00/08.0144 which is operated by Masaryk University, Czech Republic.

\textbf{Funding:} GMK is supported by the Royal Society as a Royal Society University
Research Fellow. JER is funded as a Future Faculty Leaders Fellow at the
Harvard-Smithsonian Center for Astrophysics. Early work on KELT-North was supported by
NASA Grant NNG04GO70G
